---
title: "Temperance and Governors"
execute: 
  echo: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(gt)
library(primer.data)
library(tidymodels)
library(broom)
library(marginaleffects)

# Filter the data to make the assumptions of stability and unconfoundedness more reasonable.

x <- governors |>
  mutate(election_result = ifelse(win_margin > 0, "Win", "Lose")) |>
  filter(abs(win_margin) <= 5, year > 1945)

# Fit a model. This is our data generating mechanism.

fit_causal <- linear_reg() |>
  fit(lived_after ~ election_result + win_margin + election_age + party, data = x)
```

### Background

2) Imagine you are a researcher. You want to know if winning candidates live longer. 

```{r}
tribble(
  ~`Candidate`, ~`Years Lived (Lose)`, ~`Years Lived (Win)`, ~`Election Result`, ~`Age`, ~`Win Margin`, ~`Party`, ~`Sex`, ~`Other`,
  "Joe Smith",        "18*",   "23",   "Win",  "56", "7.2",  "Republican", "Male", "...",
  "David Jones",     "22",    "28*",  "Lose", "48", "-3.5",  "Democrat",   "Male", "...",
  "...",               "..." , "..." , "..." , "..." , "..." ,  "..." , "..." , "..."
) |>
  gt() |>
  tab_header(title = "Preceptor Table") |>
  tab_spanner(label = "Potential Outcomes", columns = c(`Years Lived (Lose)`, `Years Lived (Win)`)) |>
  tab_spanner(label = "Treatment", columns = c(`Election Result`)) |>
  tab_spanner(label = "Covariates", columns = c(`Age`, `Win Margin`, `Party`, `Sex`, `Other`)) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`Candidate`)) |>
  fmt_markdown(columns = everything()) |>
  tab_footnote(
    footnote = md("Each row is a candidate running for election in 2025."),
    locations = cells_column_labels(columns = `Candidate`)
  ) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome that is not observed."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Years lived: Years lived after election, for both possible election results."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Other = additional covariates not listed here."),
    locations = cells_column_labels(columns = `Other`)
  )
```

```{r}
tribble(
  ~`Source`, ~`Candidate`, ~`Year`, ~`Years Lived (Lose)`, ~`Years Lived (Win)`, ~`Election Result`, ~`Win Margin`, ~`Age`, ~`Party`, ~`Sex`, ~`Other`,
   "...", "...", "...", "...", "...", "...", "...", "...", "...", "...", "...",
   "Data", "Earl Warren",  "1946", "18*", "23", "Win", "7.2", "56",  "Republican", "Male", "...",
   "Data", "George Wallace", "1946", "22", "28*", "Lose", "-3.5", "48", "Democrat", "Male", "...",
   "Data", "Nelson Rockefeller", "1946", "14*", "17", "Win", "9.0", "54",  "Republican", "Male", "...",
   "...", "...", "...", "...", "...", "...", "...", "...", "...", "...", "...", 
   "Preceptor Table", "Joe Smith",  "2025",      "18*",   "23",   "Win",  "56", "7.2",  "Republican", "Male", "...",
     "Preceptor Table",  "David Jones",  "2025",    "22",    "28*",  "Lose", "48", "-3.5",  "Democrat",   "Male", "...",
   "...", "...", "...", "...", "...", "...", "...", "...", "...", "...",  "..."
) |>
  gt() |>
  tab_header(title = "Population Table") |>
  tab_spanner(label = "Unit/Time", columns = c(`Candidate`, `Year`)) |>
  tab_spanner(label = "Potential Outcomes", columns = c(`Years Lived (Lose)`, `Years Lived (Win)`)) |>
  tab_spanner(label = "Treatment", columns = c(`Election Result`)) |>
  tab_spanner(label = "Covariates", columns = c(`Win Margin`, `Age`,  `Party`, `Sex`, `Other`)) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`Source`)) |>
  fmt_markdown(columns = everything()) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome that is not observed."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Years lived: Number of years lived after election for both possible outcomes."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Other = additional covariates not listed here. Age is at election."),
    locations = cells_column_spanners(spanners = "Covariates")
  )
```

## DGM

* Take a look at our fitted model. Use `tidy()` to examine the confidence intervals of the parameters. Are they any changes we might consider making?

```{r}
fit_causal |> tidy(conf.int = TRUE)
```

We can ignore partyThird party variable as its confidence intervals passes zero.

* Use AI to create a nice looking table of the parameter values. (Only work on this question if your group is going very fast and has extra time.)

```{r}
library(dplyr)
library(broom)
library(gt)

# Tidy the model and create the table
fit_causal |> 
  tidy(conf.int = TRUE) |> 
  mutate(
    p.value = formatC(p.value, format = "e", digits = 2),  # scientific notation for p-values
    estimate = round(estimate, 2),
    std.error = round(std.error, 2),
    statistic = round(statistic, 2),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2)
  ) |> 
  gt() |> 
  tab_header(
    title = "Model Parameter Estimates"
  ) |> 
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. Error",
    statistic = "t Statistic",
    p.value = "p-value",
    conf.low = "95% CI (Lower)",
    conf.high = "95% CI (Upper)"
  ) |> 
  fmt_markdown(columns = vars(term)) |> 
  tab_options(
    table.font.size = "small",
    column_labels.font.weight = "bold",
    heading.title.font.size = 16
  )

```

* Use AI to create LaTeX code for our DGM. (Only work on this question if your group is going very fast and has extra time.)

$$
\text{lived\_after}_i =
67.5647
+ 8.6345 \cdot \text{election\_resultWin}_i
- 1.4598 \cdot \text{win\_margin}_i
- 0.8869 \cdot \text{election\_age}_i
+ 4.0386 \cdot \text{partyRepublican}_i
- 9.5035 \cdot \text{partyThirdParty}_i
+ \varepsilon_i
$$

## Questions and Answers

Recall our original question: Do winning candidates live longer? Well, do they? And do other candidate characteristics matter?

* First, let's look at individual variables which are included in our DGM. Hint: Use `plot_predictions()`

```{r}
fit_causal |> predictions()
```

```{r}
fit_causal |> plot_predictions(condition = c("election_result"))
```

The plot demonstrates that winning candidates live longer.

```{r}
fit_causal |> plot_predictions(condition = c("party"))
```

Republicans live longer and data is scattered for Thirdparty.

```{r}
fit_causal |> plot_predictions(condition = c("election_age"))
```

Election age and lived after have a negative realtionship.

* Second, combine several variables --- and think about which ones you should select --- into a nice looking plot. There is no wrong answer! But some plots look better than others.

```{r}
fit_causal |> plot_predictions(condition = c("election_result", "election_age", "party"))
```



* If there is time, use the `draw = FALSE` option to `plot_predictions()` to pull out the raw data. Then, tell an AI about that tibble and ask for some nice ggplot code. Pay special attention to the subtitle.

```{r}
fit_causal |> plot_predictions(condition = c("election_result", "election_age", "party"), draw = FALSE)
```

```{r}
library(ggplot2)
library(dplyr)

# Ensure election_age is numeric
preds <- fit_causal |> 
  plot_predictions(condition = c("election_result", "election_age", "party"), draw = FALSE) |>
  mutate(
    election_age = as.numeric(election_age),  # Ensure age is numeric
    party = factor(party),
    election_result = factor(election_result)
  )

# Plot predictions
ggplot(preds, aes(x = election_age, y = estimate, color = party)) +
  geom_line(aes(linetype = election_result), size = 1) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high, fill = party),
    alpha = 0.2, color = NA
  ) +
  labs(
    title = "Predicted Lifespan After Election",
    subtitle = "Model predictions by candidate party, age, and win/loss outcome",
    x = "Candidate Age at Time of Election",
    y = "Predicted Years Lived After Election",
    color = "Party",
    fill = "Party",
    linetype = "Election Result"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
ggsave("lifespan_predictions.png", width = 8, height = 6, dpi = 300, bg = "white")

```

## Humility

* What is, truly, your best guess as to the causal effect of winning an election on longevity?

Republicans live longer and  younger ages like 32 years and wining the elction.
